% Detailed Derivations: Classical Models as Limiting Cases of Hamiltonian VFE
% For Sociology Manuscript Submission
% Author: Chris Dennis
% Date: December 2025

\documentclass[11pt]{article}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{booktabs}

\geometry{margin=1in}

\newtheorem{proposition}{Proposition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\KL}{\text{KL}}
\newcommand{\tr}{\text{tr}}

\title{Classical Social Influence Models as Limiting Cases of \\
Hamiltonian Variational Free Energy: \\
Detailed Derivations}

\author{Robert C. Dennis}
\date{December 2025}

\begin{document}

\maketitle

\begin{abstract}
We provide detailed mathematical derivations showing that six major models from sociology, psychology, and network science emerge as special or limiting cases of the Hamiltonian Variational Free Energy framework. These include DeGroot's social learning model, Friedkin-Johnsen opinion dynamics, bounded confidence models, confirmation bias, Social Impact Theory, and echo chamber formation. For each model, we present rigorous proofs or solid approximations with explicit limiting conditions, demonstrating that these seemingly disparate theories represent different dynamical regimes of a unified information-geometric framework.
\end{abstract}

\tableofcontents

\section{Introduction}

Social influence research has produced numerous models describing how individuals update beliefs through social interaction. While empirically successful in their respective domains, these models appear theoretically disconnected. We demonstrate that this fragmentation is illusory: major classical models emerge as limiting cases of a unified framework based on variational free energy minimization on statistical manifolds.

This unification proceeds by analogy to statistical mechanics. Just as thermodynamic phases (solid, liquid, gas) emerge from molecular dynamics under different conditions (temperature, pressure), we show that classical social influence models emerge from information-geometric dynamics under different conditions (friction, uncertainty, attention temperature). The underlying dynamics are always governed by natural gradient descent on the Fisher-Rao manifold, but observable behavior depends on regime parameters.

\subsection{Mathematical Foundations}

\subsubsection{The Variational Free Energy Functional}

Consider $N$ agents, each maintaining a belief distribution $q_i(x) = \N(\mu_i, \Sigma_i)$ and prior $p_i(x) = \N(\mu_{p,i}, \Sigma_{p,i})$ over a latent state $x \in \R^K$. The total variational free energy is:

\begin{align}
F[q, p] &= \sum_i \alpha \int \chi_i(c) \KL(q_i \| p_i) \, dc \label{eq:vfe_self} \\
&\quad + \sum_{i,j} \lambda_\beta \int \chi_{ij}(c) \beta_{ij}(c) \KL(q_i \| \Omega_{ij}[q_j]) \, dc \label{eq:vfe_belief} \\
&\quad + \sum_{i,j} \lambda_\gamma \int \chi_{ij}(c) \gamma_{ij}(c) \KL(p_i \| \Omega_{ij}[p_j]) \, dc \label{eq:vfe_prior} \\
&\quad - \sum_i \lambda_{\text{obs}} \int \chi_i(c) \mathbb{E}_q[\log p(o|x)] \, dc \label{eq:vfe_obs}
\end{align}

Each term serves a distinct functional role in belief dynamics. The self-coupling term \eqref{eq:vfe_self} penalizes deviation from prior expectations, with the strength parameter $\alpha$ controlling resistance to belief change; this captures the cognitive cost of abandoning established world models. The belief alignment term \eqref{eq:vfe_belief} encourages agents to align their current beliefs with those of their neighbors, where the attention weights $\beta_{ij}(c)$ are computed dynamically via softmax (detailed below) rather than fixed exogenously. The prior alignment term \eqref{eq:vfe_prior} operates analogously on priors themselves, capturing slower cultural or normative alignment with the softmax-based weights $\gamma_{ij}(c)$. Finally, the observation term \eqref{eq:vfe_obs} grounds beliefs in sensory evidence by penalizing distributions that fail to explain observed data $o$.

The spatial integration weights $\chi_i(c) \in [0,1]$ represent agent $i$'s epistemic support---the regions of state space where they maintain active beliefs---and $\chi_{ij} = \chi_i \cdot \chi_j$ captures the overlap between agents' cognitive domains.

\subsubsection{Softmax Attention Mechanism}

The belief alignment weights are computed dynamically as:
\begin{equation}
\beta_{ij}(c) = \frac{\exp\left(-\KL(q_i(c) \| \Omega_{ij}[q_j(c)]) / \kappa_\beta\right)}{\sum_k \exp\left(-\KL(q_i(c) \| \Omega_{ik}[q_k(c)]) / \kappa_\beta\right)}
\label{eq:softmax_attention}
\end{equation}

This softmax creates \emph{homophilic attention}: agents with similar beliefs (low KL divergence) receive high attention, while dissimilar agents are ignored. The temperature parameter $\kappa_\beta > 0$ controls sharpness: $\kappa_\beta \to 0$ gives winner-take-all attention, while $\kappa_\beta \to \infty$ gives uniform attention.

The transport operator $\Omega_{ij} \in SO(K)$ accounts for gauge transformations (different reference frames between agents). For flat manifolds with shared coordinates, $\Omega_{ij} = I$.

\subsubsection{Fisher-Rao Mass Matrix (Epistemic Inertia)}

The natural gradient descent dynamics on the statistical manifold are governed by the inverse Fisher metric (mass matrix):
\begin{equation}
M_i(\theta) = \Sigma_{p,i}^{-1} + \Sigma_{o,i}^{-1} + \sum_j \beta_{ij}(\theta) \, \Omega_{ij} \Sigma_{q,j}^{-1} \Omega_{ij}^T + \sum_j \beta_{ji}(\theta) \, \Sigma_{q,i}^{-1}
\label{eq:mass_matrix}
\end{equation}

This matrix comprises four physically interpretable contributions. The first term, $\Sigma_{p,i}^{-1}$, represents \emph{prior mass}---the resistance to belief change arising from established expectations, where high prior precision (low $\Sigma_p$) creates correspondingly high inertia. The second term, $\Sigma_{o,i}^{-1}$, captures \emph{observation mass}, reflecting how precise sensory evidence anchors beliefs against social pressure; agents with strong perceptual grounding are less susceptible to influence. The third term, $\sum_j \beta_{ij} \Omega_{ij} \Sigma_{q,j}^{-1} \Omega_{ij}^T$, constitutes the \emph{incoming social mass}---the accumulated inertia from attending to confident neighbors, where being pulled toward precise sources paradoxically increases one's own resistance to further change. Most notably, the fourth term, $\sum_j \beta_{ji} \Sigma_{q,i}^{-1}$, represents \emph{outgoing social mass}: the recoil from exerting influence on others. Agents with many followers (high $\sum_j \beta_{ji}$) acquire epistemic rigidity not through irrationality but through geometric necessity---the ``burden of leadership'' that renders opinion leaders resistant to revision.

\subsubsection{Dynamics: Overdamped vs. Hamiltonian Regimes}

The framework admits two dynamical regimes:

\paragraph{Overdamped (gradient flow):} When friction $\gamma \to \infty$, momentum dissipates instantly:
\begin{equation}
\frac{d\mu_i}{dt} = -M_i^{-1} \nabla_{\mu_i} F
\label{eq:overdamped}
\end{equation}

This is \emph{natural gradient descent} on the statistical manifold. The mass matrix $M^{-1}$ appears even in the overdamped limit, distinguishing natural from standard Euclidean gradient descent.

\paragraph{Hamiltonian (underdamped):} When friction $\gamma \to 0$, the system exhibits inertial dynamics:
\begin{align}
\frac{d\mu_i}{dt} &= M_i^{-1} \pi_{\mu,i} \label{eq:hamiltonian_position} \\
\frac{d\pi_{\mu,i}}{dt} &= -\nabla_{\mu_i} F - \Gamma_{ijk} \pi^j \pi^k - \gamma \pi_{\mu,i} \label{eq:hamiltonian_momentum}
\end{align}

Here $\pi_{\mu,i}$ is conjugate momentum, and $\Gamma_{ijk}$ are Christoffel symbols (geodesic corrections from metric curvature). This regime allows belief oscillations, overshooting, and non-monotonic convergence.

Classical models correspond to the \emph{overdamped regime} with various limiting conditions on parameters $(\alpha, \lambda_\beta, \kappa_\beta, \Sigma_i)$. The Hamiltonian regime generates novel predictions beyond classical theories.

\section{Derivation 1: DeGroot Social Learning (Rigorous)}

\subsection{Classical Formulation}

DeGroot's model (1974) describes social learning as iterative averaging of neighbors' beliefs:
\begin{equation}
x_i(t+1) = \sum_j w_{ij} x_j(t)
\label{eq:degroot_classical}
\end{equation}

where $W = [w_{ij}]$ is a row-stochastic matrix ($\sum_j w_{ij} = 1$) representing social influence weights. Under mild conditions, beliefs converge to a consensus determined by the network structure.

\subsection{Sociological Context}

DeGroot's model captures the fundamental sociological insight that individuals update beliefs by averaging opinions from their social network. It has been applied to jury deliberation, scientific consensus formation, and organizational decision-making. However, the model treats influence weights $w_{ij}$ as exogenous and fixed, providing no mechanism for how attention emerges from belief similarity.

\subsection{Derivation from VFE Framework}

\begin{proposition}[DeGroot as VFE Limit]
The DeGroot update rule \eqref{eq:degroot_classical} emerges from the VFE framework under the following limiting conditions:
\begin{enumerate}[label=(\roman*)]
\item Overdamped dynamics: $\gamma \to \infty$
\item Low uncertainty: $\Sigma_i \to \sigma^2 I$ with $\sigma^2$ small
\item Flat manifold: $\Omega_{ij} = I$ (shared reference frames)
\item No self-coupling: $\alpha = 0$
\item No prior alignment: $\lambda_\gamma = 0$
\item No observations: $\lambda_{\text{obs}} = 0$
\item Fixed attention: $\beta_{ij} = w_{ij}$ (constant, not softmax)
\end{enumerate}
\end{proposition}

\begin{proof}
\textbf{Step 1: Simplify VFE.}
Under conditions (iii)-(vii), the free energy \eqref{eq:vfe_belief} reduces to:
\begin{equation}
F[\mu] = \lambda_\beta \sum_{i,j} w_{ij} \int \KL(\N(\mu_i, \sigma^2 I) \| \N(\mu_j, \sigma^2 I)) \, dc
\end{equation}

For Gaussian distributions with equal covariances, the KL divergence is:
\begin{equation}
\KL(\N(\mu_i, \sigma^2 I) \| \N(\mu_j, \sigma^2 I)) = \frac{\|\mu_i - \mu_j\|^2}{2\sigma^2}
\end{equation}

Thus:
\begin{equation}
F[\mu] = \frac{\lambda_\beta}{2\sigma^2} \sum_{i,j} w_{ij} \|\mu_i - \mu_j\|^2
\label{eq:degroot_vfe_simplified}
\end{equation}

\textbf{Step 2: Compute gradient.}
Taking the gradient with respect to $\mu_i$:
\begin{align}
\nabla_{\mu_i} F &= \frac{\lambda_\beta}{\sigma^2} \sum_j w_{ij} (\mu_i - \mu_j) \\
&= \frac{\lambda_\beta}{\sigma^2} \left[\mu_i \sum_j w_{ij} - \sum_j w_{ij} \mu_j\right] \\
&= \frac{\lambda_\beta}{\sigma^2} \left[\mu_i - \sum_j w_{ij} \mu_j\right] \quad \text{(using row-stochasticity)}
\label{eq:degroot_gradient}
\end{align}

\textbf{Step 3: Apply natural gradient flow.}
In the low-uncertainty limit (condition ii), the mass matrix becomes approximately:
\begin{equation}
M_i \approx \frac{1}{\sigma^2} I
\end{equation}

Natural gradient descent \eqref{eq:overdamped} gives:
\begin{align}
\frac{d\mu_i}{dt} &= -M_i^{-1} \nabla_{\mu_i} F \\
&= -\sigma^2 I \cdot \frac{\lambda_\beta}{\sigma^2} \left(\mu_i - \sum_j w_{ij} \mu_j\right) \\
&= -\lambda_\beta \left(\mu_i - \sum_j w_{ij} \mu_j\right)
\label{eq:degroot_continuous}
\end{align}

\textbf{Step 4: Discretize.}
Applying forward Euler integration with time step $\Delta t = 1/\lambda_\beta$:
\begin{align}
\mu_i(t + \Delta t) &= \mu_i(t) + \Delta t \cdot \frac{d\mu_i}{dt} \\
&= \mu_i(t) - \lambda_\beta \Delta t \left(\mu_i - \sum_j w_{ij} \mu_j\right) \\
&= \mu_i(t) - \left(\mu_i - \sum_j w_{ij} \mu_j\right) \\
&= \sum_j w_{ij} \mu_j(t)
\end{align}

This is exactly the DeGroot update \eqref{eq:degroot_classical}. \qed
\end{proof}

\subsection{What the Unified Framework Adds}

While DeGroot's model emerges cleanly in the overdamped limit with fixed weights, the full VFE framework provides several extensions with sociological significance:

\paragraph{Dynamic attention.} Removing condition (vii) and using softmax attention \eqref{eq:softmax_attention}, influence weights become endogenous:
\begin{equation}
\beta_{ij}(t) = \frac{\exp(-\|\mu_i(t) - \mu_j(t)\|^2 / (2\sigma^2\kappa_\beta))}{\sum_k \exp(-\|\mu_i(t) - \mu_k(t)\|^2 / (2\sigma^2\kappa_\beta))}
\end{equation}

Agents pay more attention to similar others, creating homophily as an emergent property rather than assumption.

\paragraph{Uncertainty dynamics.} Removing condition (ii), beliefs are full distributions $q_i = \N(\mu_i, \Sigma_i)$. Uncertainty can increase or decrease over time, capturing phenomena like pluralistic ignorance or confidence polarization that mean-only models miss.

\paragraph{Epistemic inertia.} When agents receive asymmetric attention (some have many followers), the outgoing social mass term in \eqref{eq:mass_matrix} becomes significant:
\begin{equation}
M_i = \Sigma_p^{-1} + \Sigma_o^{-1} + \sum_j \beta_{ij} \tilde{\Sigma}_{q,j}^{-1} + \sum_j \beta_{ji} \Sigma_{q,i}^{-1}
\end{equation}

where $\tilde{\Sigma}_{q,j}^{-1} = \Omega_{ij}\Sigma_{q,j}^{-1}\Omega_{ij}^T$ is the transported precision. High-attention agents (influencers, opinion leaders) develop higher mass via the $\sum_j \beta_{ji}$ term, making their beliefs more resistant to change---a mechanistic explanation for rigidity in positions of authority.

\paragraph{Underdamped regime.} Removing condition (i) and allowing low friction $\gamma < \gamma_c$, beliefs can overshoot equilibrium and oscillate. This provides a potential mechanism for opinion cycling and instability in rapid social media discourse.

\subsection{Novel Predictions}

The unified framework generates several testable predictions beyond DeGroot's original model. First, influence weights $\beta_{ij}(t)$ should evolve dynamically as beliefs converge, a prediction testable via longitudinal network analysis tracking communication patterns alongside opinion change. Second, agents occupying central network positions---those with many followers---should update their beliefs more slowly than peripheral agents, even when controlling for initial belief extremity; this heterogeneous convergence follows from the asymmetric mass accumulation in \eqref{eq:mass_matrix}. Third, as beliefs align through social learning, agents should report increased subjective confidence (manifested as decreasing $\Sigma_i$), a prediction requiring measurement of confidence intervals rather than point estimates alone.

\section{Derivation 2: Friedkin-Johnsen Opinion Dynamics (Rigorous)}

\subsection{Classical Formulation}

Friedkin and Johnsen (1990) extended DeGroot by introducing ``stubbornness''---attachment to initial opinions:
\begin{equation}
x_i(t+1) = \alpha_i x_i(0) + (1 - \alpha_i) \sum_j w_{ij} x_j(t)
\label{eq:fj_classical}
\end{equation}

where $\alpha_i \in [0,1]$ represents agent $i$'s resistance to social influence. This model better captures polarization and persistent disagreement, as stubborn agents prevent full consensus.

\subsection{Sociological Context}

The Friedkin-Johnsen model addresses a key empirical puzzle: social groups often fail to reach consensus despite dense communication. The stubbornness parameter $\alpha_i$ is typically interpreted as a personality trait or ideological commitment. However, this raises the question: what determines stubbornness, and can it change over time?

\subsection{Derivation from VFE Framework}

\begin{proposition}[Friedkin-Johnsen as VFE Equilibrium]
The Friedkin-Johnsen equilibrium opinions emerge from the VFE framework under DeGroot conditions (i)-(vii) plus:
\begin{enumerate}[label=(\roman*), start=8]
\item Non-zero self-coupling: $\alpha > 0$
\item Fixed priors: $p_i = \N(\mu_i(0), \Sigma_p)$ (initial beliefs)
\end{enumerate}

Moreover, the stubbornness parameter $\alpha_i$ is not exogenous but emerges from prior precision and social context.
\end{proposition}

\begin{proof}
\textbf{Step 1: VFE with self-coupling.}
Including the self-coupling term \eqref{eq:vfe_self}:
\begin{align}
F[\mu] &= \alpha \sum_i \KL(\N(\mu_i, \sigma^2 I) \| \N(\mu_i(0), \Sigma_p)) \\
&\quad + \frac{\lambda_\beta}{2\sigma^2} \sum_{i,j} w_{ij} \|\mu_i - \mu_j\|^2
\end{align}

For small $\sigma^2$, the self-coupling KL divergence approximates:
\begin{equation}
\KL(\N(\mu_i, \sigma^2 I) \| \N(\mu_i(0), \Sigma_p)) \approx \frac{\|\mu_i - \mu_i(0)\|^2}{2\Sigma_p}
\end{equation}

Thus:
\begin{equation}
F[\mu] = \frac{\alpha}{2\Sigma_p} \sum_i \|\mu_i - \mu_i(0)\|^2 + \frac{\lambda_\beta}{2\sigma^2} \sum_{i,j} w_{ij} \|\mu_i - \mu_j\|^2
\label{eq:fj_vfe}
\end{equation}

\textbf{Step 2: Gradient.}
\begin{equation}
\nabla_{\mu_i} F = \frac{\alpha}{\Sigma_p}(\mu_i - \mu_i(0)) + \frac{\lambda_\beta}{\sigma^2}\left(\mu_i - \sum_j w_{ij} \mu_j\right)
\end{equation}

\textbf{Step 3: Natural gradient flow with mass $M_i = \Sigma_p^{-1}$.}
\begin{align}
\frac{d\mu_i}{dt} &= -\Sigma_p \nabla_{\mu_i} F \\
&= -\alpha(\mu_i - \mu_i(0)) - \frac{\lambda_\beta \Sigma_p}{\sigma^2}\left(\mu_i - \sum_j w_{ij} \mu_j\right)
\end{align}

\textbf{Step 4: Equilibrium solution.}
At steady state, $d\mu_i/dt = 0$:
\begin{align}
\alpha(\mu_i - \mu_i(0)) &= -\frac{\lambda_\beta \Sigma_p}{\sigma^2}\left(\mu_i - \sum_j w_{ij} \mu_j\right) \\
\mu_i \left[\alpha + \frac{\lambda_\beta \Sigma_p}{\sigma^2}\right] &= \alpha \mu_i(0) + \frac{\lambda_\beta \Sigma_p}{\sigma^2} \sum_j w_{ij} \mu_j
\end{align}

Dividing both sides by the bracketed term:
\begin{equation}
\mu_i = \frac{\alpha}{\alpha + \lambda_\beta \Sigma_p / \sigma^2} \mu_i(0) + \frac{\lambda_\beta \Sigma_p / \sigma^2}{\alpha + \lambda_\beta \Sigma_p / \sigma^2} \sum_j w_{ij} \mu_j
\end{equation}

Define the \emph{emergent stubbornness}:
\begin{equation}
\alpha_i' = \frac{\alpha}{\alpha + \lambda_\beta \Sigma_p / \sigma^2 \cdot \sum_j w_{ij}}
\label{eq:emergent_stubbornness}
\end{equation}

Then:
\begin{equation}
\mu_i = \alpha_i' \mu_i(0) + (1 - \alpha_i') \sum_j w_{ij} \mu_j
\end{equation}

which matches the Friedkin-Johnsen form \eqref{eq:fj_classical}. \qed
\end{proof}

\subsection{Mechanistic Stubbornness}

The key sociological insight is that stubbornness $\alpha_i'$ in \eqref{eq:emergent_stubbornness} is \emph{not} a fixed personality trait but emerges from two sources:

\paragraph{Prior precision $\Sigma_p^{-1}$.} Agents with strong initial convictions (low $\Sigma_p$) exhibit high stubbornness. This captures ideological commitment or expertise: a climate scientist has high prior precision about global warming, making them resistant to contrarian social influence.

\paragraph{Social coupling strength $\lambda_\beta \sum_j w_{ij}$.} Agents experiencing intense social pressure (many influential neighbors) become \emph{less} stubborn in equilibrium. However, as we show next, this same pressure increases their \emph{inertial mass}, slowing their rate of approach to equilibrium---a subtle but important distinction.

\subsection{Dynamic Stubbornness and Social Context}

Unlike the classical model where $\alpha_i$ is constant, the VFE framework predicts that stubbornness changes with social context. Consider two agents with identical priors but different network positions. Agent A, embedded in a dense network with $\sum_j w_{Aj}$ large, exhibits lower equilibrium stubbornness $\alpha_A'$ due to intense social pressure, yet paradoxically develops higher inertial mass $M_A$ from the attention received. Agent B, occupying a peripheral position with $\sum_j w_{Bj}$ small, maintains higher equilibrium stubbornness $\alpha_B'$ but lower inertial mass $M_B$. This dissociation between equilibrium position and dynamic responsiveness generates the testable prediction that centrally-located agents should express opinions closer to the group mean (low $\alpha'$) while simultaneously updating more slowly over time (high $M$).

\subsection{Novel Predictions}

The framework yields three principal predictions for Friedkin-Johnsen dynamics. Network position effects should manifest as a correlation between stubbornness and network peripherality, controlling for prior beliefs. The distinction between equilibrium displacement (stubbornness) and update rate (inertia) should be empirically separable in time-series data, as these quantities derive from different terms in the framework. Perhaps most importantly, the same individual should exhibit different degrees of stubbornness across different social contexts---for instance, between professional and personal networks---contradicting trait-based theories that treat resistance to influence as a stable personality characteristic.

\section{Derivation 3: Echo Chambers and Polarization (Rigorous)}

\subsection{Phenomenon}

Echo chambers constitute a self-reinforcing dynamical process with four characteristic stages. Initially, individuals preferentially attend to similar others---a phenomenon termed homophily. This selective attention causes in-group beliefs to converge through assimilation, while simultaneously causing out-group beliefs to diverge through polarization. The process culminates in isolation, as cross-group communication declines to negligible levels. This phenomenon has become central to understanding political polarization, radicalization, and the epistemic consequences of algorithmically-mediated social media.

\subsection{Classical Approaches}

Traditional models either \emph{assume} homophily (similarity-based edge formation) or impose exogenous group structure. The VFE framework shows that both homophily and polarization \emph{emerge endogenously} from the attention mechanism without additional assumptions.

\subsection{Derivation from VFE Framework}

\begin{proposition}[Emergent Homophily and Polarization]
Softmax attention \eqref{eq:softmax_attention} automatically creates homophilic coupling. When initial belief distributions are multimodal, this leads to stable polarized states with within-group consensus and cross-group divergence.
\end{proposition}

\begin{proof}[Proof Sketch]
\textbf{Step 1: Softmax creates homophily.}
For Gaussian beliefs with common covariance $\Sigma = \sigma^2 I$:
\begin{equation}
\beta_{ij} = \frac{\exp(-\|\mu_i - \mu_j\|^2 / (2\sigma^2 \kappa_\beta))}{\sum_k \exp(-\|\mu_i - \mu_k\|^2 / (2\sigma^2 \kappa_\beta))}
\end{equation}

Similar beliefs (small $\|\mu_i - \mu_j\|$) yield high $\beta_{ij}$ (strong attention). Dissimilar beliefs (large $\|\mu_i - \mu_j\|$) yield low $\beta_{ij}$ (ignore out-group).

\textbf{Step 2: Positive feedback loop.}
The natural gradient flow is:
\begin{equation}
\frac{d\mu_i}{dt} \propto \sum_j \beta_{ij}(\mu) (\mu_j - \mu_i)
\end{equation}

This creates a positive feedback loop through the following mechanism. Agents with similar beliefs attend preferentially to each other, generating high $\beta_{ij}$. This mutual attention causes their beliefs to converge further, $\mu_i \to \mu_j$, which in turn increases attention still more as KL divergence decreases. The process continues until cross-group attention effectively vanishes, $\beta_{ij}^{\text{cross}} \to 0$, leaving the population fragmented into internally cohesive but mutually isolated clusters.

\textbf{Step 3: Stability of polarized states.}
Consider two groups $A$ and $B$ with means $\mu_A$ and $\mu_B$. Within-group attention is:
\begin{equation}
\beta_{ij} \approx \frac{1}{|A|} \quad \text{for } i, j \in A
\end{equation}

Cross-group attention is approximately:
\begin{equation}
\beta_{ij} \approx 0 \quad \text{for } i \in A, j \in B \quad \text{when } \|\mu_A - \mu_B\| \gg \sigma\sqrt{\kappa_\beta}
\end{equation}

The system decouples into two independent subsystems, each converging internally:
\begin{equation}
\frac{d\mu_i}{dt} \approx \sum_{j \in \text{group}(i)} \beta_{ij} (\mu_j - \mu_i)
\end{equation}

\textbf{Step 4: Critical distance for polarization.}
Polarized state $\{\mu_A, \mu_B\}$ is stable when cross-group attention is negligible:
\begin{equation}
\exp\left(-\frac{\|\mu_A - \mu_B\|^2}{2\sigma^2 \kappa_\beta}\right) \ll 1
\end{equation}

This occurs when:
\begin{equation}
\|\mu_A - \mu_B\|^2 > 2\sigma^2 \kappa_\beta \log N
\label{eq:polarization_threshold}
\end{equation}

where $N$ is the number of agents. \qed
\end{proof}

\subsection{Phase Transition in Polarization}

The stability condition \eqref{eq:polarization_threshold} reveals a \emph{phase transition} in the temperature parameter $\kappa_\beta$:

\paragraph{High temperature ($\kappa_\beta$ large):} Attention is diffuse, cross-group communication persists, system converges to global consensus.

\paragraph{Low temperature ($\kappa_\beta$ small):} Attention is sharp, cross-group communication collapses, system locks into polarized state.

The critical temperature scales with the initial belief separation:
\begin{equation}
\kappa_\beta^{\text{crit}} \sim \frac{\|\mu_A(0) - \mu_B(0)\|^2}{2\sigma^2 \log N}
\end{equation}

This provides a quantitative prediction: polarization is more likely when (i) initial disagreement is large, (ii) uncertainty is low, or (iii) attention is selective (low $\kappa_\beta$).

\subsection{Connection to Filter Bubbles}

Social media platforms that use engagement-based ranking effectively lower $\kappa_\beta$ (sharpen attention toward similar content). The VFE framework predicts this design choice should increase polarization, consistent with empirical observations.

Interventions to reduce polarization should target $\kappa_\beta$: increasing exposure diversity (raising temperature) or increasing epistemic humility (raising uncertainty $\sigma^2$) can prevent polarization phase transition.

\subsection{Novel Predictions}

The framework generates three principal predictions concerning echo chamber dynamics. First, varying attention selectivity experimentally should reveal a sharp phase transition: polarization should emerge abruptly at a critical value of $\kappa_\beta$ rather than gradually, analogous to thermodynamic phase transitions. Second, polarized states should be reversible under appropriate interventions---raising $\kappa_\beta$ through increased cross-group exposure, or increasing epistemic uncertainty $\sigma^2$ through epistemic humility interventions, can in principle dissolve echo chambers by restoring cross-group attention. Third, smaller groups should polarize more easily than larger ones, as the threshold in \eqref{eq:polarization_threshold} scales with $\log N$; this asymmetric polarization may explain why small online communities often exhibit more extreme views than larger populations.

\section{Derivation 4: Bounded Confidence Models (Solid Approximation)}

\subsection{Classical Formulation}

Hegselmann-Krause (2002) and Deffuant et al. (2000) introduced bounded confidence: agents only interact with others within a threshold distance $\epsilon$:
\begin{equation}
x_i(t+1) = \begin{cases}
\text{avg}\{x_j(t) : |x_j(t) - x_i(t)| < \epsilon\} & \text{if } |\{j : |x_j - x_i| < \epsilon\}| > 0 \\
x_i(t) & \text{otherwise}
\end{cases}
\label{eq:hk_classical}
\end{equation}

This generates clustering: opinions fragment into groups separated by more than $\epsilon$, with consensus within each group.

\subsection{Sociological Context}

Bounded confidence models capture the idea that individuals ignore opinions too far from their own---a form of selective exposure or cognitive dissonance avoidance. The threshold $\epsilon$ is typically treated as an exogenous parameter, but sociologically, it should depend on context.

\subsection{Derivation from VFE Framework}

\begin{proposition}[Bounded Confidence as Low-Temperature Limit]
The bounded confidence dynamics approximate the VFE framework in the low-temperature regime $\kappa_\beta \to 0$, with effective threshold:
\begin{equation}
\epsilon_{\text{eff}} \approx \sigma\sqrt{2\kappa_\beta \log N}
\end{equation}
\end{proposition}

\begin{proof}[Proof Sketch]
\textbf{Step 1: Low-temperature softmax.}
As $\kappa_\beta \to 0$, the softmax \eqref{eq:softmax_attention} becomes increasingly sharp:
\begin{equation}
\beta_{ij} = \frac{\exp(-\|\mu_i - \mu_j\|^2 / (2\sigma^2\kappa_\beta))}{\sum_k \exp(-\|\mu_i - \mu_k\|^2 / (2\sigma^2\kappa_\beta))}
\end{equation}

Let $d_{ij} = \|\mu_i - \mu_j\|$. For $d_{ij}$ significantly larger than the minimum distance, the exponential becomes negligible:
\begin{equation}
\exp(-d_{ij}^2 / (2\sigma^2\kappa_\beta)) \approx 0 \quad \text{when } d_{ij}^2 \gg 2\sigma^2\kappa_\beta \log N
\end{equation}

\textbf{Step 2: Effective threshold.}
Agents within the effective radius receive substantial attention:
\begin{equation}
\epsilon_{\text{eff}} = \sigma\sqrt{2\kappa_\beta \log N}
\end{equation}

Beyond this radius, attention decays exponentially to zero.

\textbf{Step 3: Approximate dynamics.}
The natural gradient flow becomes:
\begin{align}
\frac{d\mu_i}{dt} &\propto \sum_j \beta_{ij} (\mu_j - \mu_i) \\
&\approx \sum_{j : d_{ij} < \epsilon_{\text{eff}}} \frac{1}{|N_i(\epsilon)|} (\mu_j - \mu_i) \\
&= \text{avg}\{\mu_j - \mu_i : \|\mu_j - \mu_i\| < \epsilon_{\text{eff}}\}
\end{align}

which matches the Hegselmann-Krause update (in continuous time). \qed
\end{proof}

\subsection{Key Difference: Soft vs. Hard Threshold}

A crucial distinction separates the VFE framework from classical bounded confidence models. Hegselmann-Krause employs a hard cutoff, $\beta_{ij} = 1/|N_i|$ if $d < \epsilon$ and zero otherwise, creating a discontinuous attention function. The VFE framework instead produces a soft threshold via the exponential form $\beta_{ij} = \exp(-d^2 / (2\sigma^2\kappa_\beta)) / Z_i$, which decays smoothly with distance.

This distinction carries two significant implications. Regarding realism, actual social attention likely exhibits smooth falloff rather than abrupt cutoffs; people do not entirely ignore slightly-too-distant opinions but rather attend to them with diminishing weight. The soft threshold better captures this psychological reality. Regarding mathematical tractability, soft thresholds are differentiable everywhere, enabling analytic study of stability, bifurcations, and phase transitions that would be inaccessible with discontinuous dynamics.

\subsection{Adaptive Threshold}

Unlike fixed-$\epsilon$ models, the effective threshold in the VFE framework depends dynamically on multiple parameters through the relationship $\epsilon_{\text{eff}} = f(\sigma, \kappa_\beta, N)$. This functional dependence generates several novel predictions. The uncertainty effect predicts that higher epistemic uncertainty $\sigma$ increases tolerance for distant opinions (larger effective $\epsilon$), consistent with empirical findings that epistemic humility reduces polarization. The group size effect predicts that larger groups exhibit wider effective thresholds due to the $\log N$ scaling, potentially explaining why small online echo chambers often display more extreme views than larger populations. Finally, context-dependence implies that the same individual should exhibit different tolerance thresholds across different social contexts characterized by varying $\kappa_\beta$---a person might engage with distant opinions in academic settings (high $\kappa_\beta$) while ignoring them in partisan political contexts (low $\kappa_\beta$).

\subsection{Novel Predictions}

The bounded confidence derivation yields three testable predictions. Rather than sharp clustering boundaries, the soft threshold predicts gradual attention decay measurable via interaction frequencies across opinion distances. Experimental manipulation of $\kappa_\beta$---for instance, through platform design choices that encourage or discourage engagement with diverse content---should produce predictable changes in effective tolerance thresholds. Interventions that induce epistemic humility, operationalized as raising subjective uncertainty $\sigma$, should measurably increase cross-group engagement by widening the effective confidence bound.

\section{Derivation 5: Confirmation Bias (Requires Natural Gradient)}

\subsection{Phenomenon}

Confirmation bias---the tendency to update beliefs less from counter-attitudinal evidence than from supportive evidence---is one of the most robust findings in social psychology. It has been alternately explained as motivated reasoning, cognitive dissonance reduction, or Bayesian conservatism.

\subsection{Classical Approaches}

Most models treat confirmation bias as an \emph{ad hoc} deviation from rational Bayesian updating, requiring additional parameters (e.g., directional weighting of evidence). The VFE framework shows that asymmetric updating emerges naturally from the geometry of the statistical manifold.

\subsection{Derivation from VFE Framework}

\begin{proposition}[Confirmation Bias from Epistemic Inertia]
In the natural gradient flow regime, agents with high prior precision or many followers exhibit confirmation bias without any non-Bayesian mechanisms.
\end{proposition}

\begin{proof}[Proof Sketch]
\textbf{Step 1: Natural gradient with mass matrix.}
The update dynamics are:
\begin{equation}
\frac{d\mu_i}{dt} = -M_i^{-1} \nabla_{\mu_i} F
\end{equation}

where the full mass matrix is:
\begin{equation}
M_i = \Sigma_{p,i}^{-1} + \Sigma_{o,i}^{-1} + \sum_j \beta_{ij} \tilde{\Sigma}_{q,j}^{-1} + \sum_j \beta_{ji} \Sigma_{q,i}^{-1}
\end{equation}

with $\tilde{\Sigma}_{q,j}^{-1} = \Omega_{ij}\Sigma_{q,j}^{-1}\Omega_{ij}^T$ the transported neighbor precision.


\textbf{Step 2: Effect of high prior precision.}
When prior precision $\Sigma_p^{-1}$ is large (agent is confident in prior):
\begin{equation}
M_i^{-1} \approx (\Sigma_p^{-1})^{-1} = \Sigma_p \quad (\text{small})
\end{equation}

For the same force $\nabla F$, the velocity is small:
\begin{equation}
\frac{d\mu}{dt} = \Sigma_p \nabla F \ll \nabla F \quad \text{when } \Sigma_p \text{ is small}
\end{equation}

\textbf{Step 3: Effect of social mass.}
When $\sum_j \beta_{ji} \Sigma_{q,i}^{-1}$ is large (many followers paying attention to agent $i$):
\begin{equation}
M_i \text{ large} \implies M_i^{-1} \text{ small} \implies \frac{d\mu_i}{dt} \text{ small}
\end{equation}

Note the crucial distinction: $\beta_{ji}$ (others attending to $i$) contributes to $i$'s mass, not $\beta_{ij}$ (whom $i$ attends to).


\textbf{Step 4: Asymmetric response to evidence.}
Consider evidence $o$ that contradicts agent $i$'s belief. The observation term contributes:
\begin{equation}
\nabla_{\mu_i} F_{\text{obs}} \propto (\mu_i - \mu_{\text{evidence}})
\end{equation}

Update magnitude:
\begin{equation}
\Delta \mu_i \propto M_i^{-1} (\mu_i - \mu_{\text{evidence}}) = \left[\Sigma_p^{-1} + \Sigma_o^{-1} + \sum_j \beta_{ij} \tilde{\Sigma}_{q,j}^{-1} + \sum_j \beta_{ji} \Sigma_{q,i}^{-1}\right]^{-1} (\mu_i - \mu_{\text{evidence}})
\end{equation}

The magnitude of belief updates thus depends inversely on the total epistemic mass. Agents with high prior precision (low $\Sigma_p$) develop large mass and correspondingly small updates. Those with strong sensory grounding (low $\Sigma_o$) remain anchored against social pressure. Agents with many followers (large $\sum_j \beta_{ji}$) accumulate substantial outgoing social mass, rendering them resistant to new information. Even agents who merely attend to confident neighbors (large $\sum_j \beta_{ij}\tilde{\Sigma}_{q,j}^{-1}$) develop increased incoming social mass. Through these four mechanisms, confirmation bias emerges naturally: beliefs become resistant to contradictory evidence not through motivated reasoning but through the geometric structure of the statistical manifold. \qed
\end{proof}

\subsection{Critical Assumption: Natural Gradient}

This derivation requires interpreting the dynamics as \emph{natural gradient descent} (flow on the statistical manifold) rather than standard Euclidean gradient descent. The mass matrix $M^{-1}$ appears only in the natural gradient formulation.

\paragraph{Justification.} Natural gradient descent is the unique geometrically-principled learning rule on statistical manifolds, invariant under reparameterization (Amari, 1998). The Fisher-Rao metric $M$ is the Riemannian metric on the space of probability distributions. Standard gradient descent is only appropriate for Euclidean spaces; beliefs are distributions, not vectors.

\paragraph{Status.} This derivation is solid \emph{if} we accept natural gradient as the correct dynamics for learning on manifolds---a standard assumption in information geometry.

\subsection{Social Component of Confirmation Bias}

The novel insight of this derivation concerns the \emph{outgoing social mass} term $\sum_j \beta_{ji} \Sigma_{q,i}^{-1}$---representing the recoil from influencing others. This term generates three distinctive predictions. First, confirmation bias should be audience-dependent: the same individual should exhibit stronger resistance to counter-evidence when they have many followers (as with public intellectuals compared to private citizens), because $\sum_j \beta_{ji}$ increases with audience size. Second, the effect is self-precision-weighted, scaling with agent $i$'s own confidence $\Sigma_{q,i}^{-1}$; confident leaders therefore experience more rigidity than uncertain ones with identical follower counts. Third, the effect is dynamic: as attention structure changes---for instance, when content goes viral---confirmation bias should increase in real-time through the evolving $\beta_{ji}$ terms.

\subsection{Novel Predictions}

The confirmation bias derivation yields three experimentally tractable predictions. The follower effect predicts that manipulating perceived audience size should produce corresponding changes in resistance to counter-evidence. The precision effect, somewhat counterintuitively, predicts that counter-evidence from highly confident sources may be \emph{more} strongly resisted if it triggers defensive responses that increase the target's own precision. The temporal dynamics prediction suggests that sudden attention spikes---as when content goes viral---should produce temporarily elevated confirmation bias during the attention peak, with subsequent relaxation as attention disperses.

\section{Derivation 6: Social Impact Theory (Interpretive Mapping)}

\subsection{Classical Formulation}

Latané's Social Impact Theory (1981) posits that social influence is a multiplicative function of three factors:
\begin{equation}
\text{Impact} = f(\text{Strength} \times \text{Immediacy} \times \text{Number})
\label{eq:sit_classical}
\end{equation}

Here strength encompasses the expertise, status, and power of the influence source; immediacy captures proximity in space or time; and number reflects how many sources are simultaneously present. This qualitative principle has been widely applied across social psychology but has historically lacked a quantitative mechanistic formulation that would render it precisely testable.

\subsection{Mapping to VFE Framework}

The mass matrix \eqref{eq:mass_matrix} provides a natural quantitative interpretation of Latané's factors:
\begin{equation}
M_i = \Sigma_p^{-1} + \Sigma_o^{-1} + \sum_j \beta_{ij} \Omega_{ij} \Sigma_{q,j}^{-1} \Omega_{ij}^T + \sum_j \beta_{ji} \Sigma_{q,i}^{-1}
\end{equation}

\paragraph{Strength $\leftrightarrow$ $\Sigma_{q,j}^{-1}$.} Source precision---operationalized as confidence or expertise---contributes directly to the mass experienced by the target. Experts possess high precision (large $\Sigma_q^{-1}$) and consequently make large contributions to target mass, while uncertain sources with low precision contribute negligibly. This provides a mechanistic account of why expert opinions carry more weight: they literally increase the epistemic inertia of those who attend to them.

\paragraph{Immediacy $\leftrightarrow$ Transport penalty $\|\Omega_{ij} - I\|$.} Spatial and temporal distance enter the framework through gauge transformations between reference frames. Agents who are ``close'' in the relevant sense have nearly aligned frames ($\Omega_{ij} \approx I$), incurring small KL penalties and receiving high attention weights $\beta_{ij}$. Distant agents, whose frames are significantly rotated, incur large KL penalties and receive correspondingly low attention. The immediacy function thus emerges naturally from the softmax mechanism:
\begin{equation}
f_{\text{imm}}(\Omega_{ij}) = \exp(-\|\Omega_{ij} - I\|_F^2 / \kappa_\beta)
\end{equation}

\paragraph{Number $\leftrightarrow$ $\sum_j$.} More sources → more terms in sum → larger social mass.

\subsection{Quantitative Formulation}

The social mass contribution from source $j$ is:
\begin{equation}
\Delta M_i^{(j)} = \beta_{ij} \Omega_{ij} \Sigma_{q,j}^{-1} \Omega_{ij}^T
\end{equation}

In scalar form (for simplicity):
\begin{equation}
\Delta M_i^{(j)} \approx (\text{Number: }1) \times (\text{Strength: }\Sigma_{q,j}^{-1}) \times (\text{Immediacy: }f(\Omega_{ij}))
\end{equation}

Total impact:
\begin{equation}
M_i \approx \Sigma_p^{-1} + \Sigma_o^{-1} + \sum_j (\text{Strength}_j \times \text{Immediacy}_{ij} \times \beta_{ij}) + \sum_j (\text{Self-precision} \times \beta_{ji})
\label{eq:sit_vfe}
\end{equation}

where the final term captures "recoil" whereby having followers increases one's own inertia.


\subsection{What the VFE Framework Adds}

\paragraph{Exact quantitative formula.} Latané's principle is qualitative; \eqref{eq:sit_vfe} gives precise predictions testable via measurement of $\Sigma_q$, $\Omega$, and $\beta$.

\paragraph{Time-varying impact.} As beliefs and attention evolve, mass changes dynamically:
\begin{equation}
M_i(t) = \Sigma_p^{-1} + \Sigma_o^{-1} + \sum_j \beta_{ij}(t) \tilde{\Sigma}_{q,j}(t)^{-1} + \sum_j \beta_{ji}(t) \Sigma_{q,i}(t)^{-1}
\end{equation}

This captures how influence waxes and wanes with changing confidence, attention, and audience structure.

\paragraph{Asymmetry.} Social impact is not reciprocal:
\begin{equation}
\Delta M_i^{(j)} \neq \Delta M_j^{(i)}
\end{equation}

High-status sources influence low-status targets more than the reverse, even with symmetric communication.

\paragraph{Testability.} Each component of the VFE formulation admits independent measurement: strength through confidence ratings and expertise assessments, immediacy through physical distance, temporal lags, and measures of frame alignment, and number through network degree and audience size metrics. The framework becomes testable by measuring these components separately and then verifying whether their combined effect matches the quantitative predictions of \eqref{eq:sit_vfe}.

\subsection{Caveat: Interpretive Mapping}

This is an \emph{interpretive correspondence}, not a formal equivalence. Latané's formula is intentionally vague (``multiplicative function of''); the VFE framework provides one specific quantitative instantiation. Other interpretations are possible, though \eqref{eq:sit_vfe} has the virtue of being uniquely derived from information geometry.

\subsection{Novel Predictions}

The Social Impact Theory mapping generates three distinctive predictions. The confidence modulation prediction states that impact should scale with source confidence $\Sigma_q^{-1}$, testable by experimentally manipulating perceived expertise while holding other factors constant. The frame alignment prediction addresses the phenomenon of ``talking past each other'': sources operating in different conceptual frameworks (high $\|\Omega - I\|$) should exert reduced impact even when physically proximate, suggesting that shared language and conceptual alignment matter independently of spatial proximity. The temporal decay prediction states that impact should decline with communication lag according to a specific functional form, with decay rate determined by the attention temperature $\kappa_\beta$.

\section{Derivation 7: Diffusion of Innovations (Solid Approximation)}

\subsection{Classical Formulation}

Rogers (1962) identified a characteristic S-curve pattern in the adoption of innovations across populations, with distinct adopter categories emerging from the distribution of adoption times. Innovators (approximately 2.5\% of the population) adopt first, followed by early adopters (13.5\%), early majority (34\%), late majority (34\%), and finally laggards (16\%). The cumulative adoption curve follows a logistic function:
\begin{equation}
N(t) = \frac{K}{1 + e^{-r(t - t_0)}}
\label{eq:scurve}
\end{equation}
where $K$ is the carrying capacity, $r$ the growth rate, and $t_0$ the inflection point. Classical models treat adoption thresholds as fixed individual characteristics, leaving unexplained why the distribution takes this particular form and what determines the critical mass required for takeoff.

\subsection{Derivation from VFE Framework}

\begin{proposition}[S-Curve from Attention Dynamics]
The logistic adoption curve emerges from the VFE framework when agents decide between two belief states (adopt/reject) based on social attention from prior adopters, with heterogeneous prior precisions determining adoption order.
\end{proposition}

\begin{proof}[Proof Sketch]
\textbf{Step 1: Binary belief dynamics.}
Consider agents with beliefs $\mu_i \in \{0, 1\}$ representing rejection and adoption respectively. The social force toward adoption is proportional to attention-weighted adopter fraction:
\begin{equation}
F_i^{\text{adopt}} = \sum_{j: \mu_j = 1} \beta_{ij}
\end{equation}

Adoption occurs when this force exceeds the agent's resistance threshold, which is determined by their prior precision $\Sigma_{p,i}^{-1}$ anchoring them to the status quo.

\textbf{Step 2: Threshold heterogeneity.}
If prior precisions are distributed across the population according to some distribution $\rho(\Sigma_p^{-1})$, agents with low precision (innovators) adopt when $F^{\text{adopt}}$ is small, while high-precision agents (laggards) require nearly unanimous adoption. For a log-normal distribution of precisions---empirically reasonable given multiplicative accumulation of confidence---the resulting adoption curve is approximately logistic.

\textbf{Step 3: Positive feedback and critical mass.}
The attention mechanism creates positive feedback. As more agents adopt, non-adopters receive increased social pressure through the $\beta_{ij}$ terms. Initially, with few adopters, the attention-weighted signal is weak and only the lowest-precision agents adopt. Once adoption crosses a critical threshold, the attention signal becomes strong enough to cascade through the remaining population. The critical mass occurs when:
\begin{equation}
\sum_{j: \mu_j = 1} \beta_{ij} > \Sigma_{p,\text{median}}^{-1} \cdot c
\end{equation}
for typical agents, where $c$ is a coupling constant. \qed
\end{proof}

\subsection{Adopter Categories as Mass Distribution}

The Rogers categories emerge naturally from the distribution of epistemic mass $M_i$. Innovators possess low mass (uncertain priors, few social connections anchoring them to the status quo), enabling rapid adoption even against social consensus. Early adopters have moderate mass but occupy network positions that expose them to innovators' attention. The early majority waits until the attention-weighted adopter signal exceeds their threshold. The late majority, with higher mass from both prior precision and social embeddedness, requires near-saturation. Laggards possess the highest mass---whether from extreme prior certainty, isolation from adopter networks, or both---and adopt only under overwhelming social pressure or not at all.

A key prediction distinguishes this framework from classical diffusion models: adopter category membership should be context-dependent rather than a stable individual trait. An innovator in technology adoption (low precision regarding technical matters) might be a laggard in fashion (high precision regarding aesthetic standards). The same agent exhibits different effective mass across domains.

\subsection{The Commitment Trap}

The outgoing social mass term $\sum_j \beta_{ji} \Sigma_{q,i}^{-1}$ generates a novel prediction regarding innovation abandonment. Early adopters who influence others accumulate social mass from those attention connections, making them increasingly resistant to abandoning the innovation even if problems emerge. Opinion leaders become trapped by their followers: the very act of evangelizing an innovation creates epistemic rigidity that prevents acknowledging its failures. This explains the common observation that early advocates are often the last to admit an innovation has failed, not through motivated reasoning but through geometric necessity.

\subsection{Novel Predictions}

The diffusion derivation yields three distinctive predictions. First, adoption thresholds should correlate with measurable prior precision and network centrality, with the interaction between these factors determining adopter category more reliably than either alone. Second, opinion leaders who actively promote innovations should exhibit slower abandonment rates than passive early adopters, controlling for initial enthusiasm---the commitment trap prediction. Third, the critical mass threshold should depend systematically on attention temperature $\kappa_\beta$: highly connected networks with uniform attention (high $\kappa_\beta$) should require smaller critical mass than sparse networks with selective attention (low $\kappa_\beta$).

\section{Derivation 8: Collective Intentionality and We-Intentions (Speculative)}

\subsection{Phenomenon}

Collective intentionality---the capacity for shared intentions, joint attention, and ``we-mode'' cognition---has been extensively analyzed by philosophers including Searle (1990), Bratman (1993), and Tuomela (2005). The central puzzle is that ``we intend to $X$'' appears irreducible to collections of individual intentions ``I intend that we $X$.'' Agents in we-mode share not just aligned goals but a distinct mode of cognition characterized by mutual responsiveness, collective commitment, and group identification. The transition from I-mode to we-mode represents a qualitative shift in social organization that existing models struggle to formalize mathematically.

\subsection{Framework Mapping}

The VFE framework suggests a geometric interpretation of collective intentionality through the interplay of gauge alignment, attention structure, and prior convergence. While this mapping remains speculative, it offers a principled mathematical structure for what has traditionally been a purely philosophical distinction.

\paragraph{Gauge frame alignment.} Individual agents maintain reference frames $\phi_i$ determining how they interpret information. When agents interact superficially, their frames may differ substantially ($\|\Omega_{ij} - I\|$ large), requiring costly translation at each communication. The prior alignment term $\text{KL}(p_i \| \Omega_{ij}[p_j])$ encourages frame convergence over time. We-mode may correspond to a phase where frames have aligned sufficiently ($\Omega_{ij} \approx I$ for in-group members) that agents share a common reference frame for interpreting the world.

\paragraph{Attention density transition.} In I-mode, attention is sparse and transactional---agents attend to specific others for specific purposes. In we-mode, attention becomes dense and persistent---agents continuously monitor group members' beliefs and coordinate their updates. Mathematically, this corresponds to a transition from low to high effective coupling $\lambda_\beta$. Above a critical coupling strength, individual belief dynamics become entrained, and the group behaves as a coherent unit.

\paragraph{Prior convergence.} The deepest aspect of collective intentionality may be shared priors---not merely aligned current beliefs but shared expectations, values, and interpretive frameworks. When the prior alignment term dominates ($\lambda_\gamma$ large), agents' generative models converge, creating genuinely shared cognition rather than merely coordinated individual cognition.

\subsection{Phase Transition Interpretation}

\begin{proposition}[We-Mode as Phase Transition]
The transition from I-mode to we-mode may correspond to a phase transition in the VFE system, occurring when coupling strength exceeds a critical threshold and gauge frames align.
\end{proposition}

Consider a group of $N$ agents with pairwise coupling. In the disordered (I-mode) phase, agents maintain distinct frames and update beliefs relatively independently. The order parameter---perhaps the average frame alignment $\langle \|\Omega_{ij} - I\| \rangle$---remains large. As coupling strength increases, a critical point is reached where frames spontaneously align, attention becomes mutually entrained, and the system enters an ordered (we-mode) phase characterized by collective dynamics.

This interpretation explains several features of collective intentionality. The apparent irreducibility of ``we intend'' to ``I intend'' reflects the emergent nature of collective order---just as ferromagnetism cannot be reduced to individual spins, we-mode cannot be reduced to I-mode intentions. The special normative force of collective commitments reflects the epistemic mass accumulated through mutual attention---defection would require overcoming substantial social inertia. The fragility of we-mode under defection reflects proximity to the phase boundary: a single agent reverting to I-mode can trigger a cascade of frame misalignment.

\subsection{Group Identification}

The framework offers a geometric account of group identification. When an agent adopts the group's prior as their own ($p_i \approx p_{\text{group}}$), they inherit the group's inertial mass. The self-coupling term $\alpha \text{KL}(q_i \| p_i)$ then anchors the agent's beliefs to group expectations. Identification is not merely cognitive labeling but geometric restructuring: the agent's belief dynamics become constrained by group-level attractors.

This predicts that strongly identified group members should exhibit slower belief updating in response to individual evidence (higher effective mass from prior coupling), faster updating in response to perceived group consensus (lower barrier to alignment with group prior), and resistance to information that threatens group integrity (the prior itself is at stake, not just current beliefs).

\subsection{Caveat: Speculative Status}

This derivation is the most speculative in the present work. Collective intentionality involves normative and phenomenological dimensions that may resist purely geometric treatment. The framework offers a mathematical vocabulary for describing we-mode but does not explain how or why conscious agents experience collective intentions as they do. Nevertheless, the mapping suggests empirical predictions---phase transition signatures, coupling thresholds, frame alignment measurements---that could lend precision to philosophical debates.

\subsection{Novel Predictions}

Despite its speculative status, the we-intentions mapping generates three testable predictions. First, groups should exhibit measurable frame alignment (operationalized perhaps through semantic similarity in language use or coordination in ambiguous tasks) that increases with interaction density and predicts collective action success. Second, the transition to we-mode should exhibit phase transition signatures: hysteresis (groups that have achieved we-mode should maintain it at coupling levels below the threshold for initial formation), critical slowing (dynamics near the transition should be slower than far from it), and susceptibility peaks (small perturbations should have large effects near the critical point). Third, group identification should correlate with measurable changes in belief updating rates, with strongly identified members showing the mass-enhancement pattern described above.

\section{Derivation 9: Kuhnian Paradigm Shifts (Solid/Interpretive)}

\subsection{Classical Formulation}

Kuhn's \emph{Structure of Scientific Revolutions} (1962) identified a pattern in scientific change that challenged cumulative progress narratives. Science alternates between \emph{normal science}---puzzle-solving within an accepted paradigm---and \emph{revolutionary science}---paradigm competition and replacement. Paradigms provide shared exemplars, methods, and interpretive frameworks that enable efficient research but also create blind spots. Anomalies accumulate until a crisis triggers revolutionary upheaval. Crucially, paradigms are \emph{incommensurable}: scientists working in different paradigms may literally see different things, use the same terms with different meanings, and lack neutral ground for adjudication.

\subsection{Mapping to VFE Framework}

The VFE framework provides a natural geometric interpretation of Kuhnian dynamics. Paradigms correspond to metastable states in belief space---local minima of the free energy landscape separated by barriers. Normal science is overdamped relaxation within a basin of attraction. Revolution is transition between basins, requiring sufficient energy to cross the barrier.

\paragraph{Paradigms as shared priors and frames.}
A scientific paradigm comprises shared priors $p_{\text{paradigm}}$ (what counts as a good explanation, plausible mechanism, relevant evidence) and shared gauge frames $\Omega_{\text{paradigm}}$ (how concepts map to observations). Scientists within a paradigm share these structures, enabling efficient communication ($\text{KL}$ penalties are small for within-paradigm interaction) and cumulative progress (new results integrate smoothly with existing beliefs).

\paragraph{Incommensurability as gauge mismatch.}
Scientists in different paradigms operate with different frames. The gauge transformation $\Omega_{AB}$ between paradigm $A$ and paradigm $B$ may be far from identity, creating substantial translation costs:
\begin{equation}
\text{KL}(q_A \| \Omega_{AB}[q_B]) \gg \text{KL}(q_A \| q_{A'})
\end{equation}
for within-paradigm neighbors $A'$. This geometric distance, not mere disagreement about facts, constitutes incommensurability. Scientists in different paradigms may use identical words (``mass,'' ``gene,'' ``species'') with systematically different meanings, corresponding to different frame rotations applied to shared underlying structure.

\paragraph{Normal science as overdamped dynamics.}
During normal science, the community operates near equilibrium within a paradigm's basin. Beliefs update incrementally in response to new evidence, following approximately gradient flow:
\begin{equation}
\frac{d\mu_i}{dt} = -M_i^{-1} \nabla_{\mu_i} F
\end{equation}
Anomalies---observations poorly explained by the paradigm---increase free energy locally but are typically absorbed through auxiliary adjustments rather than paradigm revision. The prior term $\text{KL}(q \| p_{\text{paradigm}})$ creates resistance to fundamental revision.

\paragraph{Crisis and instability.}
Anomalies accumulate as free energy, but the paradigm's basin structure initially contains them. Crisis occurs when accumulated free energy exceeds the barrier height:
\begin{equation}
F_{\text{current}} > F_{\text{barrier}} = \min_{\text{path}} \max_{\text{along path}} F
\end{equation}
At this point, the system becomes unstable with respect to paradigm-crossing fluctuations. In the underdamped regime (low friction, young scientists, rapid communication), the system may spontaneously tunnel to an alternative basin.

\paragraph{Revolution as basin transition.}
Paradigm change corresponds to a collective transition from one basin to another. This requires both an alternative basin (a competing paradigm with lower free energy for the accumulated evidence) and sufficient activation energy to cross the barrier. The mass matrix explains why revolutions are generational: established scientists have high mass (from prior precision and accumulated social attention), making barrier-crossing exponentially unlikely. Young scientists have low mass, enabling exploration of alternative basins.

\subsection{Planck's Principle}

The framework provides a mechanistic account of Planck's observation that ``science advances one funeral at a time.'' Senior scientists accumulate epistemic mass through multiple channels. Their extensive training creates high prior precision $\Sigma_p^{-1}$ for the paradigm's assumptions. Their influence over students, grants, and publications generates substantial outgoing social mass $\sum_j \beta_{ji} \Sigma_{q,i}^{-1}$. Their position atop attention hierarchies means their beliefs shape the field's direction.

This accumulated mass creates a barrier to paradigm change that operates through legitimate epistemic mechanisms rather than irrationality. A senior scientist's resistance to a new paradigm reflects genuine difficulty in revising deeply held priors, real costs of abandoning frameworks that have proven productive, and the weight of responsibilities to students and colleagues working within the existing paradigm. The claim is not that senior scientists are \emph{wrong} to resist but that the geometry of their epistemic position makes resistance inevitable.

Young scientists, by contrast, have low mass: limited prior investment in the old paradigm, few followers whose beliefs they must consider, and positions in the attention network that expose them to frontier developments. They can explore alternative basins that would be inaccessible to high-mass established researchers.

\subsection{Incommensurability and Translation}

The gauge structure provides a precise account of partial incommensurability. Complete incommensurability ($\Omega_{AB}$ arbitrary, no translation possible) is rare; complete commensurability ($\Omega_{AB} = I$, translation trivial) means the paradigms are not genuinely different. Typical cases involve substantial but structured gauge rotations, where translation is possible but costly and imperfect.

This explains why paradigm debates often involve talking past each other: participants are applying different gauge transformations to map concepts onto shared observations. It also explains why some translations succeed (the gauge group has structure that permits certain rotations) while others fail (some conceptual reorganizations are geometrically inaccessible).

\subsection{Novel Predictions}

The Kuhnian derivation yields three distinctive predictions. First, paradigm change rates should correlate with the age-mass distribution of the field: fields dominated by high-mass senior scientists should exhibit slower paradigm change than fields with flatter hierarchies, controlling for evidential pressure. Second, interdisciplinary work should accelerate paradigm change by introducing low-gauge-distance concepts from adjacent fields, effectively providing ``stepping stones'' across the free energy barrier. Third, the framework predicts \emph{hysteresis} in paradigm dynamics: the evidence threshold required to trigger a revolution should be higher than the threshold required to prevent reversion to the old paradigm, since barrier-crossing involves asymmetric costs depending on which basin the community currently occupies.

\section{Summary Table: Derivation Quality}

\begin{table}[h]
\centering
\begin{tabular}{@{}lccp{5.5cm}@{}}
\toprule
\textbf{Model} & \textbf{Rigor} & \textbf{Status} & \textbf{Notes} \\
\midrule
DeGroot & $\checkmark\checkmark\checkmark$ & Rigorous & Exact limit with QED proof \\
Friedkin-Johnsen & $\checkmark\checkmark\checkmark$ & Rigorous & Exact limit; emergent stubbornness \\
Echo Chambers & $\checkmark\checkmark\checkmark$ & Rigorous & Direct consequence of softmax \\
Bounded Confidence & $\checkmark\checkmark$ & Solid & Soft approximation of hard threshold \\
Confirmation Bias & $\checkmark\checkmark$ & Solid & Requires natural gradient assumption \\
Social Impact Theory & $\checkmark\checkmark$ & Interpretive & Qualitative correspondence \\
Diffusion of Innovations & $\checkmark\checkmark$ & Solid & S-curve from attention dynamics \\
Collective Intentionality & $\checkmark$ & Speculative & Phase transition interpretation \\
Kuhnian Paradigm Shifts & $\checkmark\checkmark$ & Solid/Interp. & Metastable basins; gauge incommens. \\
\bottomrule
\end{tabular}
\caption{Quality assessment of each derivation. Rigor levels: $\checkmark\checkmark\checkmark$ = Rigorous (formal proof), $\checkmark\checkmark$ = Solid (strong approximation or additional assumptions), $\checkmark$ = Speculative (preliminary).}
\label{tab:rigor}
\end{table}

\section{Novel Predictions Summary}

Beyond recovering classical models, the VFE framework generates novel empirical predictions that distinguish it from existing theories. These predictions fall into several thematic clusters.

\paragraph{Attention and network dynamics.} The prediction of \emph{dynamic attention} states that influence networks should restructure as beliefs converge or diverge, with $\beta_{ij}(t)$ evolving in measurable ways via longitudinal network data. The \emph{epistemic inertia} prediction asserts that agents receiving high social attention should update their beliefs more slowly than peripheral agents, controlling for prior strength; this is testable via social media data tracking opinion leaders' update rates. The \emph{commitment trap} prediction from diffusion theory suggests that early adopters who actively promote innovations should exhibit slower abandonment rates than passive adopters, as influence accumulates as mass.

\paragraph{Phase transitions and collective phenomena.} The framework predicts multiple phase transitions with characteristic signatures. Polarization should emerge sharply at a critical attention selectivity $\kappa_\beta^{\text{crit}}$, exhibiting hysteresis, critical slowing, and susceptibility peaks. The transition from I-mode to we-mode in collective intentionality should exhibit analogous signatures as gauge frames align. Paradigm shifts should display hysteresis, with revolution thresholds exceeding reversion thresholds due to asymmetric barrier costs.

\paragraph{Context-dependence and domain specificity.} The prediction of \emph{context-dependent stubbornness} contradicts trait-based theories: the same individual should exhibit variable resistance to influence across contexts. Similarly, adopter categories (innovator to laggard) should be domain-specific rather than stable personality traits, reflecting context-dependent mass distributions. Scientists' resistance to paradigm change should correlate with field-specific mass accumulation, not general conservatism.

\paragraph{Geometric structure of communication.} The \emph{gauge mismatch} prediction states that communication difficulty should scale with frame rotation $\|\Omega_{ij} - I\|$, independently of belief similarity. This explains why paradigm debates involve ``talking past each other'' and why interdisciplinary translation is costly. Frame alignment should be measurable through semantic similarity in language use and should predict coordination success.

\paragraph{Generational and temporal dynamics.} Planck's principle emerges quantitatively: paradigm change rates should correlate inversely with the age-mass distribution of the field. The \emph{underdamped oscillations} prediction suggests that in low-friction environments, beliefs may overshoot equilibrium, measurable via high-frequency polling. Sudden attention spikes (viral content) should produce temporarily elevated confirmation bias during the peak.

Testing even a subset of these predictions would provide strong empirical validation of the unifying framework, while falsification of any would constrain the parameter regimes where the theory applies.

\section{Conclusion}

We have demonstrated that nine major models from sociology, psychology, philosophy, and science studies emerge as limiting cases or regimes of the Hamiltonian Variational Free Energy framework. These include DeGroot social learning, Friedkin-Johnsen opinion dynamics, bounded confidence, echo chambers, confirmation bias, Social Impact Theory, diffusion of innovations, collective intentionality, and Kuhnian paradigm shifts. Three derivations are rigorously proven (DeGroot, Friedkin-Johnsen, echo chambers), four are solid approximations (bounded confidence, confirmation bias, diffusion, Kuhn), and two are interpretive or speculative mappings (Social Impact Theory, collective intentionality).

This unification reveals that classical models differ not in fundamental mechanisms but in \emph{parameter regimes}: friction ($\gamma$), uncertainty ($\Sigma$), attention temperature ($\kappa_\beta$), self-coupling ($\alpha$), and gauge alignment ($\Omega$). The framework generates novel testable predictions beyond classical theories, particularly regarding dynamic attention, epistemic inertia from social position, phase transitions in collective phenomena, and the geometric structure of incommensurability.

The key insight is that \emph{social structure creates epistemic consequences through information geometry}. Agents receiving high social attention develop inertial mass that resists belief revision---not through irrationality or motivated reasoning, but as a geometric necessity on the statistical manifold. This provides a mechanistic foundation for phenomena ranging from confirmation bias to Planck's principle, from the commitment trap in innovation adoption to the fragility of collective intentionality.

The gauge-theoretic component of the framework opens particularly rich territory. Incommensurability between paradigms, the difficulty of interdisciplinary translation, and the costs of frame misalignment in collective action all receive geometric interpretations. The framework suggests that communication is not merely exchange of propositions but alignment of reference frames---a perspective with implications for understanding both everyday miscommunication and fundamental theoretical disagreements.

\section{Extended Framework: Meta-Agents, Renormalization, and Participatory Dynamics}

The derivations above treat agents as fixed entities interacting through belief dynamics. A richer framework---developed in detail in the companion manuscript on participatory information geometry---extends this to hierarchical multi-scale systems where agents themselves form and dissolve dynamically. This section sketches the sociological implications of four key constructions from that framework.

\subsection{Meta-Agent Formation: Institutions as Emergent Collective Entities}

When agents achieve sufficient belief coherence---measured by small KL divergence after gauge transport---they can form \emph{meta-agents}: higher-order collective entities that are not mere aggregates but autonomous dynamical systems with their own beliefs, priors, and gauge frames.

\begin{definition}[Meta-Agent Formation Condition]
A cluster of scale-$s$ agents $\{i, j, \ldots\}$ forms a scale-$(s+1)$ meta-agent when:
\begin{equation}
C_{\text{belief}} \cdot C_{\text{model}} > \Gamma_{\min}, \quad \text{where } C_{\text{belief}} = \exp\left(-\frac{1}{|S|}\sum_{i,j \in S} \text{KL}(q_i \| \Omega_{ij}[q_j])\right)
\end{equation}
and similarly for model coherence $C_{\text{model}}$.
\end{definition}

\paragraph{Sociological interpretation.} This formalizes the emergence of institutions, organizations, and collective actors. A family forms when individuals achieve sufficient alignment in beliefs about shared domestic reality. A scientific discipline forms when researchers achieve consensus on methods, exemplars, and evaluative standards. A nation forms when populations develop shared narratives, norms, and expectations. Crucially, the meta-agent is not a metaphor but a genuine dynamical entity: it evolves according to the same free energy minimization principles as individual agents, and it can influence its constituents through top-down feedback.

\paragraph{The consensus-diversity balance.} The framework predicts a fundamental tension between consensus (enabling meta-agent formation and collective action) and diversity (preventing \emph{epistemic death}---the pathological limit where all agents become informationally identical). Groupthink represents excessive consensus; fragmentation represents insufficient consensus. Healthy institutions occupy the critical regime where coherence enables collective action while diversity preserves adaptive capacity. This provides a quantitative framework for analyzing organizational pathologies.

\paragraph{Gauge covariance and organizational translation.} Different members of an organization may use different ``languages'' (gauge frames) while maintaining genuine alignment on underlying beliefs. The marketing department and engineering department may describe the same product differently, but if their beliefs are related by well-defined gauge transport $\Omega_{ij}$, they constitute a coherent meta-agent. Organizational dysfunction often corresponds to gauge mismatch: departments using incompatible frames with no well-defined translation.

\subsection{Dynamical Renormalization: Scale Transitions in Social Systems}

The formation of meta-agents implements an information-theoretic analog of renormalization group transformations. Instead of coarse-graining over physical space, we coarse-grain over \emph{belief coherence}---agents form meta-agents not because they are spatially proximate but because their beliefs align.

\paragraph{Scale-invariant dynamics.} The free energy functional at scale $(s+1)$ has the same form as at scale $s$:
\begin{equation}
F^{(s+1)} = \sum_I \text{KL}(q_I^{(s+1)} \| p_I^{(s+1)}) + \sum_{IJ} \beta_{IJ}^{(s+1)} \text{KL}(q_I \| \Omega_{IJ}[q_J]) + \cdots
\end{equation}
This scale invariance suggests that the same dynamical principles govern individuals, groups, organizations, and societies---differing only in effective coupling constants and characteristic timescales.

\paragraph{Emergent timescale hierarchy.} Meta-agents at higher scales evolve more slowly than their constituents:
\begin{equation}
\tau^{(0)} < \tau^{(1)} < \tau^{(2)} < \cdots < \tau^{(s_{\max})}
\end{equation}
This emerges from three mechanisms: spatial/informational averaging smooths out fast fluctuations, coordinated multi-agent updates require more time, and accumulated epistemic mass resists rapid change. The sociological implication is that institutional change is \emph{necessarily} slower than individual opinion change---not due to bureaucratic friction but due to the geometric structure of collective belief dynamics.

\paragraph{Emergent properties.} Meta-agents exhibit properties absent at lower scales: longer coherence lengths (organizational culture spanning an entire company), higher precision (collective knowledge exceeding any individual's), and emergent coordination patterns (market behavior not reducible to individual trades). This provides a mathematical foundation for claims about ``emergent social properties'' that have long been debated in sociological theory.

\subsection{The Ouroboros Tower: Cultural Hierarchies and Historical Memory}

A key extension allows agents to receive priors not just from their immediate parent meta-agent but from \emph{all ancestral scales} simultaneously:
\begin{align}
p_i^{(s)} &\leftarrow \Omega_{i,I_1}[q_{I_1}^{(s+1)}] \quad \text{(parent: family)} \\
h_i^{(s,1)} &\leftarrow \Omega_{i,I_2}[q_{I_2}^{(s+2)}] \quad \text{(grandparent: community)} \\
h_i^{(s,2)} &\leftarrow \Omega_{i,I_3}[q_{I_3}^{(s+3)}] \quad \text{(great-grandparent: tradition)}
\end{align}
with exponentially decaying weights $\lambda_k = \lambda_0 \cdot \gamma^k$ for $k$-generational distance.

\paragraph{Cultural heritage as hyperprior.} This formalizes how individuals are shaped not just by their immediate social context but by entire civilizational heritage. A scientist is influenced not only by their research group (scale 1) but by their discipline's paradigm (scale 2), the broader scientific tradition (scale 3), and Enlightenment epistemology (scale 4). Each ancestral level contributes a hyperprior with exponentially decaying weight, explaining why deep cultural assumptions are both influential and difficult to articulate explicitly.

\paragraph{Tradition and innovation.} The exponential decay $\gamma^k$ predicts that recent ancestry matters more than distant, but distant ancestry still exerts measurable influence. This provides a quantitative model for the interplay between tradition (high-scale hyperpriors) and innovation (low-scale belief updates). Highly traditional societies correspond to large $\gamma$ (slow decay, strong ancestral influence); rapidly modernizing societies correspond to small $\gamma$ (fast decay, weak ancestral influence).

\paragraph{Self-referential closure.} For agents at the top of the hierarchy, priors are formed by observing the entire system state---society watching itself through media, polling, collective representations. This creates Wheeler's ``self-excited circuit'': society constructs collective priors that flow down to shape individual beliefs, which aggregate into new collective representations, which society re-observes. The participatory loop has no external ground---neither individuals nor collectives are fundamental; both co-constitute through perpetual information exchange.

\subsection{Bidirectional Information Flow: Micro-Macro Dynamics}

The framework formalizes both bottom-up and top-down causation in social systems:

\paragraph{Bottom-up: emergence.} Individual beliefs aggregate into collective beliefs through gauge-covariant averaging. Public opinion is not merely the sum of individual opinions but a geometrically constructed meta-belief that accounts for frame differences between individuals. This explains why aggregate measures (polls, markets, votes) can diverge from naive averages---the aggregation procedure matters.

\paragraph{Top-down: constraint.} Meta-agent beliefs become priors for constituents through gauge-covariant transport: $p_i \leftarrow \Omega_{i,I}[q_I]$. Social norms, institutional expectations, and cultural scripts are not merely external pressures but become internalized priors that shape individual belief dynamics from within. The mechanism is not coercion but prior propagation---individuals adopt collective perspectives as their own expectations.

\paragraph{Non-equilibrium dynamics.} A functioning participatory system must remain far from equilibrium. Complete consensus (epistemic death) would eliminate information gradients and freeze evolution. The framework predicts that healthy societies maintain sufficient diversity to sustain adaptation while achieving sufficient consensus to enable collective action. This tension drives perpetual social dynamics---there is no stable end-state, only ongoing negotiation between convergence and divergence.

\subsection{Sociological Predictions from Participatory Dynamics}

The participatory framework generates several distinctive predictions for social systems:

\paragraph{Formation thresholds.} Organizational formation should exhibit threshold effects: below critical coherence $\Gamma_{\min}$, no stable meta-agent forms; above it, meta-agents emerge and persist. This predicts \emph{phase transitions} in organizational formation---gradual increases in belief alignment produce sudden emergence of collective identity.

\paragraph{Hierarchical depth and inertia.} Individuals embedded in deep institutional hierarchies should update beliefs more slowly than those in shallow hierarchies, controlling for other factors. Each ancestral scale contributes hyperpriors that increase effective mass, explaining why bureaucrats in ancient institutions resist change more than entrepreneurs in startups.

\paragraph{Self-observation effects.} Societies with more collective self-observation (pervasive media, frequent polling, social media analytics) should exhibit qualitatively different dynamics---potentially faster consensus formation but also increased susceptibility to collective delusions when self-observation feeds back into the observed system.

\paragraph{Timescale ratios.} The framework predicts specific relationships between characteristic timescales at adjacent scales: $\tau^{(s+1)}/\tau^{(s)} = f(\beta, N, \sigma)$ where $f$ depends on coupling strengths, group sizes, and uncertainty. This could be empirically tested by comparing the rate of individual opinion change to the rate of institutional norm change across different organizational contexts.

\paragraph{Gauge mismatch and organizational conflict.} The framework predicts that inter-organizational conflict should correlate with gauge mismatch $\|\Omega_{IJ} - I\|$ between organizational frames, independently of belief similarity. Two organizations with similar goals but different conceptual frameworks may conflict more intensely than organizations with opposed goals but aligned frames---because frame mismatch makes even agreement difficult to recognize.

\appendix

\section{Appendix: Technical Details}

\subsection{Fisher-Rao Metric and Natural Gradient}

For a family of distributions $p_\theta(x)$ parameterized by $\theta \in \R^d$, the Fisher-Rao metric is:
\begin{equation}
g_{ij}(\theta) = \mathbb{E}_{p_\theta}\left[\frac{\partial \log p_\theta}{\partial \theta_i} \frac{\partial \log p_\theta}{\partial \theta_j}\right]
\end{equation}

This metric is the unique (up to scaling) Riemannian metric invariant under sufficient statistics transformations.

For multivariate Gaussian $\N(\mu, \Sigma)$, the Fisher metric in the $\mu$-coordinates is $\Sigma^{-1}$, exactly the mass matrix bare term. The social coupling term extends this to multi-agent settings.

Natural gradient descent minimizes a function $F(\theta)$ along geodesics:
\begin{equation}
\frac{d\theta}{dt} = -g^{-1}(\theta) \nabla_\theta F
\end{equation}

Standard gradient descent ($g = I$) is only correct for Euclidean parameter spaces; belief distributions live on curved manifolds.

\subsection{Christoffel Symbols and Geodesic Corrections}

In the Hamiltonian regime, the momentum equation includes geodesic corrections:
\begin{equation}
\frac{d\pi_i}{dt} = -\nabla_i F - \Gamma_{ijk} \pi^j \pi^k
\end{equation}

where:
\begin{equation}
\Gamma_{ijk} = \frac{1}{2} g^{i\ell} \left(\frac{\partial g_{j\ell}}{\partial \theta_k} + \frac{\partial g_{k\ell}}{\partial \theta_j} - \frac{\partial g_{jk}}{\partial \theta_\ell}\right)
\end{equation}

These terms ensure trajectories follow geodesics on the manifold. For static metrics ($\partial g/\partial \theta = 0$), they vanish; for parameter-dependent mass matrices (our case), they create nonlinear coupling between position and momentum.

\subsection{Symplectic Integration}

Hamiltonian dynamics preserve phase space volume (Liouville's theorem) and approximately conserve energy. Standard integrators (Runge-Kutta, etc.) violate these properties, leading to artificial energy drift.

Symplectic integrators (Verlet, leapfrog, Ruth) preserve the symplectic 2-form:
\begin{equation}
\omega = \sum_i d\theta_i \wedge d\pi_i
\end{equation}

guaranteeing long-term stability and bounded energy drift. For our simulations, we use:
\begin{align}
\pi(t + \tfrac{\Delta t}{2}) &= \pi(t) + \tfrac{\Delta t}{2} f(\theta(t)) \\
\theta(t + \Delta t) &= \theta(t) + \Delta t \, g^{-1}(\theta) \pi(t + \tfrac{\Delta t}{2}) \\
\pi(t + \Delta t) &= \pi(t + \tfrac{\Delta t}{2}) + \tfrac{\Delta t}{2} f(\theta(t + \Delta t))
\end{align}

where $f = -\nabla F - \Gamma \pi \pi - \gamma \pi$.

\bibliographystyle{plain}
\bibliography{references}

\end{document}
